### 2016-05-30 text/Chapter08.tex
Edits and tweaks (Chapter 8 and 9 mostly)

The prototype implementation discussed in the previous chapter directly follows the two
coeffect calculi presented in this thesis. This links together all aspects of the coeffect
theory (type system, semantics, safety proofs) and it provides practical support for the
presented theoretical work. However, the theory of coeffects could be embedded in practical
programming languages in a various ways and the theory could be further developed in a number
of different directions.

The most important alternative directions revisit three design decisions made in this thesis.
First, rather than discussing flat and structural coeffects, we can treat them in a single
unified calculus tracking both kinds of context dependence. Second, rather than using ``language
semantics'' style (Section~\ref{sec:path-sem-langs}), we could use a ``meta-language'' style
and extend the host language with explicit constructs for working with coeffects. Third,
rather than implementing a prototype coeffect language, we could embed coeffects into an
existing language using techniques inspired by Haskell's ``do'' notation.

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start by unifying the two kinds of coeffects discussed so far. We introduce the
  \emph{unified coeffect calculus} (Section~\ref{sec:further-unified}), which generalizes flat
  and structural systems and can be instantiated to track both flat and structural properties.

@@

\noindent
The unified coeffect calculus presented in the first part of the chapter is a novel result that
completes the theory developed in this thesis. The next two sections present important future and
related work, respectively.

@@

\begin{itemize}
\item The structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
  explicitly, but they can be derived thanks to the fact that variables in the context can be
  freely reordered.

@@

\item In the flat coeffect calculus, we distinguish between \emph{splitting} context demands
  and \emph{merging} of context demands ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model appending and splitting vectors) are invertible
  and so the structural coeffect algebra uses the same tensor product $\atimes$. However, the
  $\cpar$ operation is still needed in the contraction rule.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We include explicit rules for context manipulation in the calculus; in systems that arise from flat
calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize
the structure of coeffect annotations using the notion of a ``container'' which can be specialized
to obtain a single annotation or a vector of annotations. This could potentially be used to capture
other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
between splitting and merging of context demands (using the notation $\apar$ and $\azip$,
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they differ and provide the needed flexibility.

@@

Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise
as some containers may have multiple shapes of the same length, \eg~differently balanced trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$.
Here, $S$ is a set of shapes and $P$ is a shape-indexed family $P : S \rightarrow \ident{Set}$
of positions.
\end{definition}

@@

\subsection{Structure of coeffects}
In the structural coeffect calculus, the coeffect annotation was formed by a vector of coeffect
scalars. The annotations in the unified coeffect calculus are similar, but a \emph{vector} is
replaced with a more general \emph{container}. The primitive coeffect annotations in the unified
calculus are formed of \emph{coeffect scalars}. The coeffect scalars are equipped with the same
structure as in structural coeffect calculus (Definition~\ref{def:structural-scalar}). In this
section, we refer to it as \emph{\cclrd{unified coeffect scalar}} (and we repeat the definition
below). Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect
scalar values are attached to the free-variable context. Finally, we define the
\emph{\aclrd{unified coeffect algebra}} which consists of shape-indexed coeffect scalar values.

@@

\paragraph{Coeffect scalars.}
The following definition of the coeffect scalar structure repeats the
Definition~\ref{def:structural-scalar} from the previous chapter.

\begin{definition}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \,\cseq, \,\cpar, \,\cunit, \,\czero, \,\cleq)$ is a set
$\C$ together with elements $\cunit, \,\czero \in \C$, binary operations $\cseq, \,\cpar$ such that
$(\C, \,\cseq, \,\cunit)$ and $(\C, \,\cpar, \,\czero)$ are monoids and a binary relation $\cleq$ such
that $(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}
\\
\end{array}
\end{equation*}
%
The following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
%
In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
\begin{equation*}
\begin{array}{l}
 \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
 \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
\end{array}
\end{equation*}
\end{definition}

\noindent
As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
an underlying category structure; $\cunit$ and $\czero$ represent an accessed and unused variable,
respectively.

The $\cpar$ operation models combining of context demands arising from multiple parts of a
program. The meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural coeffects) or requirements attached to the whole context of
multiple sub-exp\-ressions (flat coeffects).

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how are scalar coeffects attached to
free-variable contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed positions,
the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.

@@

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
to $\Gamma$. We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation
$\SP(\sclrd{s})$ returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in
  \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects.
We write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
The coeffect container is equipped with an operation that appends shapes (when concatenating
variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton
context.

\begin{definition}
A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$
comprises a container $\SHP$ with a binary operation $\stimes$ on $\SH$ for appending shapes, a
mapping from free-variable contexts to shapes $\slen{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
\SH$ such that $(\SH, \stimes, \sempty)$ is a monoid.

@@

\noindent
As discussed earlier, we use two kinds of coeffect containers that describe the structure of vectors
(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):

\begin{example}
\label{ex:further-struct-shape}
Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP(\sclrd{n})=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$,
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} =
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

@@

\begin{example}
\label{ex:further-flat-shape}
Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP(\sflat) = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

@@

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \aclrd{unified
coeffect algebra} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:

\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

@@

The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:further-semantics}),
where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of
a morphism that merges the capabilities provided by two contexts (in the syntactic
reading, splits the context demands), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$
is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single
context into two parts (in the syntactic reading, merges their context demands).
Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and
$\apar$ in \emph{conclusions}.

@@

\subsection{Unified coeffect type system}
\label{sec:further-types}
The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
how the unified system differs from the structural system.

The type system for the unified coeffect calculus is parameterized by a coeffect scalar
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors (without duplicate
variables) with associativity built-in. The order of variables matters, but can be changed
using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars.

\paragraph{Syntax-driven rules.}
The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$.
The (\emph{const}) rule annotates empty context with a special annotation of the shape $\sempty$.

@@

@@

\paragraph{Structural rules.}
The merging/splitting operations in the structural rules are changed in the same way as in the
rules above. It is also worth noting that structural coeffect system used vectors of multiple
elements. For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch})
rule denoted two-element vectors. In the unified system, this is replaced with merging/splitting
of two lifted scalars: $\alift{\cclrd{s}} \azip \alift{\cclrd{t}}$ and $\alift{\cclrd{t}} \apar \alift{\cclrd{s}}$.

In structural systems, the two notations mean the same thing -- we are simply concatenating or splitting two
singleton vectors. However, this generalization allows us to capture flat coeffect systems as well.
The lifting operation in flat systems simply returns the lifted scalar and operators $\azip$ and
$\apar$ correspond to operations on coeffect scalars. As discussed in Section~\ref{sec:further-flat},
thanks to the properties of coeffect scalars, contraction, weakening and exchange that do not
affect the coeffect annotation are admissible for all flat systems embedded in the unified
coeffect calculus.

@@

The unified calculus closely resembles the structural system and so using it to model structural
systems is easy -- given a coeffect scalar, we use the coeffect container that describes a \emph{vector}
of annotations (Example~\ref{ex:further-struct-shape}) and define a coeffect algebra formed by
a vector (free monoid) of scalars.

@@

Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:further-flat-shape}).
In this setting, the context annotations $\C^{ \SP(\sflat) } = \{0\} \rightarrow \C$ are equivalent
to coeffect scalars $\C$. In addition to the coeffect scalar structure, we need to define $\apar$
and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$
(merging of shaped coeffect annotations). The $\azip$ operation (corresponding to $\czip$
in flat coeffect calculus) needs to be provided explicitly.

@@

\begin{lemma}
\label{thm:further-weak}
A unified coeffect calculus capturing flat properties admits weakening that does not change the
coeffect annotation.
\end{lemma}
\begin{proof}
The rule is admissible using the following derivation:
\begin{equation*}
\inference
  { \inference
      {\coctx{\Gamma}{ \cclrd{r} } \vdash e : \tau}
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \;\apar\; \alift{ \czero }} \vdash e : \tau} }
  { \inference
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \,\cpar\, \czero} \vdash e : \tau}
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r}} \vdash e : \tau} }
\end{equation*}

@@

\begin{lemma}
A unified coeffect calculus capturing flat properties admits exchange that does not change the
coeffect annotation.
\end{lemma}
\begin{proof}
We use the idempotence of $\czip$ and $\cpar$ together with the (\emph{exch}) rule:
\begin{equation*}
\inference
  { \inference
      { \inference
          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r}} \vdash e : \tau}
          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r}} \vdash e : \tau} }
      { \coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}} \vdash e : \tau } }
  { \inference
      {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \apar \alift{\cclrd{r}} \apar \alift{\cclrd{r}} \apar \cclrd{r}} \vdash e : \tau}
      {\inference
         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}} \vdash e : \tau}
         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}} \vdash e : \tau} } }
\end{equation*}

@@

\noindent
Although the different reading does not affect formal properties of the systems, it is important
to understand the difference when discussing the two systems as they provide a different
intuition.

>>>>

The prototype implementation discussed in the previous chapter directly follows the two
coeffect calculi presented in this thesis. It links together all aspects of the coeffect
theory (type system, semantics, safety proofs) and it provides practical support for the
presented theoretical work. In this chapter, we consider alternative ways of embedding the
theory of coeffects in practical programming languages and we also outline further directions
in which the theory can be further developed.

The most important alternative directions revisit three design decisions made in this thesis.
First, rather than discussing flat and structural coeffects, we can treat them in a single
unified calculus tracking both kinds of context dependence (Section~\ref{sec:further-unified}).
Second, rather than using ``language semantics'' style (Section~\ref{sec:path-sem-langs}), we
could use a ``meta-language'' style and extend the host language with explicit constructs for
working with coeffects (Section~\ref{sec:further-meta}). Third, rather than implementing a
prototype coeffect language, we could embed coeffects into an existing language using techniques
inspired by Haskell's ``do'' notation (Section~\ref{sec:further-further}).

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start by unifying the two kinds of coeffects discussed so far. We introduce the
  \emph{unified coeffect calculus} (Section~\ref{sec:further-unified}), which generalizes flat
  and structural systems and which can be instantiated to track both flat and structural properties.

@@

\noindent
The unified coeffect calculus presented in Section~\ref{sec:further-unified} is a novel result that
completes the theory developed in this thesis and has been discussed in detail in a separate
joint publication by the author \cite{coeffects-icfp14}. We present it as further work here,
because we do not supplement the unified calculus with the same operational treatment as the flat
and structural coeffect calculi. Finally, the Sections~\ref{sec:further-meta} presents
important related work and Section~\ref{sec:further-further} presents important further work.

@@

\begin{itemize}
\item The structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not
  explicit, but they can be derived thanks to the fact that variables in the context can be
  freely reordered. The (\emph{var}) rule differs to permit implicit exchange and weakening
  in the flat system (by ignoring other variables in the context) and (\emph{app}) implicitly
  permits contraction (by duplicating the context passed to sub-expressions).

@@

\item In the flat coeffect calculus (Figure~\ref{fig:flat-types}), we distinguish between
  \emph{splitting} context demands and \emph{merging} of context demands ($\cpar$ and $\czip$,
  respectively). In the structural coeffect calculus, the operations (which model appending and
  splitting vectors) are invertible and so the structural coeffect algebra uses the same tensor
  product $\atimes$. In essence $\cpar$ and $\czip$ represent under- and over-approximation of
  $\atimes$. In the structural coeffect system, the $\cpar$ operation is still needed
  in the contraction rule.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We include explicit rules for context manipulation in the calculus; in systems that arise from flat
calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize
the structure of coeffect annotations using the notion of a ``container'' which can be specialized
to obtain a single annotation or a vector of annotations. This could potentially be alternatively
specialized to capture other structures such as trees in bunched typing \cite{substruct-bunched}.
Finally, we keep the $\cpar$ operation (needed in the contraction rule), but we distinguish
between splitting and merging of context demands (using the notation $\apar$
and $\azip$, respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they differ and provide the needed flexibility.

@@

Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
lists)\footnote{Shapes can be intuitively thought of as sizes, but this is not precise
as some containers have multiple shapes of the same length, \eg~for perfectly ballanced, left- and right-leaning trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$.
Here, $S$ is a set of shapes and $P$ is a shape-indexed family of sets such that
for each shape $s\in S$, $P_s$ (also written as $P(s)$) is a set of positions for shape $s$.
\end{definition}

@@

\subsection{Structure of coeffects}
In the structural coeffect calculus, coeffect annotations are formed by vectors of coeffect
scalars. The annotations in the unified coeffect calculus are similar, but the notion of \emph{vector}
is replaced with a more general \emph{container}. The primitive coeffect annotations in the unified
calculus are formed by coeffect scalars, which have the same structure as in the structural
coeffect calculus (Definition~\ref{def:structural-scalar}). In this section, we refer to it as
\emph{\cclrd{unified coeffect scalar}} (we repeat the definition below for clarity).
Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect
scalar values are attached to the free-variable context. Finally, we define the
\emph{\aclrd{unified coeffect algebra}} which consists of shape-indexed coeffect scalar values.

@@

\paragraph{Coeffect scalars.}
The following definition of the coeffect scalar structure repeats
Definition~\ref{def:structural-scalar} from the Chapter~\ref{ch:structural}, with the only change
that we now refer to it as \emph{\cclrd{unified coeffect scalar}}.

\begin{definition}
\label{def:unified-scal}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \,\cseq, \,\cpar, \,\cunit, \,\czero, \,\cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, binary operations $\cseq, \cpar$ such that
$(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and a binary relation $\cleq$ such
that $(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}
\\
\end{array}
\end{equation*}
%
In addition, the following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

\noindent
As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
an underlying category structure; $\cunit$ and $\czero$ represent an accessed and unused variable,
respectively. The $\cpar$ operation models combining of context demands arising from multiple parts of a
program. Its meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural coeffects) or requirements attached to the whole context of
multiple sub-exp\-ressions (flat coeffects).

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how are scalar coeffects attached to free-variable
contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed sets of positions,
the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.

@@

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
to $\Gamma$. We write $\sclrd{l} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$ and
$\SP_{\sclrd{l}}$ is the \emph{set} of positions for the given shape. A coeffect $\aclrd{\textbf{r}}$
can now be seen as a function $\SP_{\sclrd{l}} \rightarrow \C$ that returns a scalar coefect
$\cclrd{r}\in\C$ for each of the position defined by the shape $\sclrd{l}$.
We write the set of positions for a shape $\sclrd{l}$ as $\SP(\sclrd{l})$ and a function that
returns scalar coeffect as an exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{l})}$.

Additionally, the coeffect container is equipped with an operation that appends shapes (when concatenating
variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton
context.

\begin{definition}
\label{def:unified-cont}
A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$
comprises a container $\SHP$ with a binary operation $\stimes$ on $\SH$ for appending shapes, a
mapping from free-variable contexts to shapes $\slen{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
\SH$ such that $(\SH, \stimes, \sempty)$ is a monoid.

@@

\noindent
As discussed earlier, we use two kinds of coeffect containers. Those that describe the structure of vectors
(for structural coeffects) and those that describe the shape of trivial singleton container (for flat coeffects):

\begin{example}
\label{ex:further-struct-shape}
A structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP(\sclrd{n})=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$,
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} =
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

@@

\begin{example}
\label{ex:further-flat-shape}
A flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP(\sflat) = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

@@

\begin{definition}
\label{def:unified-alg}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \aclrd{unified
coeffect algebra} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\begin{array}{rcl}
\apar_{\sclrd{m},\sclrd{n}} &:& \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})} \\
\azip_{\sclrd{m},\sclrd{n}} &:& \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{array}
\end{equation*}
\end{definition}

@@

The difference between $\azip$ and $\apar$ is clarified by the semantics \cite{coeffects-icfp14},
where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of
a morphism that merges the capabilities provided by two contexts (in the syntactic
reading, splits the context demands), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$
is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single
context into two parts (in the syntactic reading, merges their context demands).
Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and
$\apar$ in \emph{conclusions}.

@@

\subsection{Unified coeffect type system}
\label{sec:further-types}
We define the unified coeffect system in Figure~\ref{fig:unif-types}. It resembles the structural
type system shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we
focus on how the unified system differs from the structural system.

The type system for the unified coeffect calculus is parameterized by a coeffect scalar
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ (Definition~\ref{def:unified-scal}) together with a
coeffect algebra $(\azip, \apar, \azero)$ (Definition~\ref{def:unified-alg})
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors of distinct
variables with associativity built-in. The order of variables matters, but can be changed
using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
are shape-indexed coeffects (rather than simple vectors as before). As before, function arrows
are annotated with coeffect scalars.

\paragraph{Syntax-driven rules.}
The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
differently. The $\alift{-}$ operation does not create a \emph{vector}, but a coeffect container
of shape $\sunit$ that returns the coeffect scalars $\cunit$ for all positions in the singleton shape.
The (\emph{const}) rule annotates empty context with a special annotation of the shape $\sempty$.

@@

~

@@

\paragraph{Structural rules.}
The merging/splitting operations in the structural rules are generalized in the same way as in the
rules above. It is also worth noting that structural coeffect systems use vectors of elements while
the unified system uses the container structure.
For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch})
rule denoted two-element vectors. In the unified system, this is replaced with merging/splitting
of two lifted scalars: $\alift{\cclrd{s}} \azip \alift{\cclrd{t}}$ and $\alift{\cclrd{t}} \apar \alift{\cclrd{s}}$.

In structural systems, the two notations mean the same thing -- we are simply concatenating or splitting two
singleton vectors. However, this generalization allows us to capture flat coeffect systems as well.
The lifting operation in flat systems simply returns the lifted scalar; operators $\azip$ and
$\apar$ correspond to operations on coeffect scalars. As discussed in Section~\ref{sec:further-flat},
thanks to the properties of coeffect scalars, the operations of contraction, weakening and exchange
that do not affect the coeffect annotation are admissible for all flat systems embedded in the
unified coeffect calculus.

@@

The unified calculus closely resembles the structural system and so using it to model structural
systems is easy -- given a (structural) coeffect scalar, we use the coeffect container that describes
a \emph{vector} of annotations (Example~\ref{ex:further-struct-shape}) and define a coeffect
algebra formed by a vector (free monoid) of scalars.

@@

Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:further-flat-shape}).
In this setting, the context annotations $\C^{ \SP(\sflat) }$ are defined as function values
$\{0\} \rightarrow \C$. The domain of the function is a singleton set and so the values are equivalent
to coeffect scalars $\C$. In addition to the coeffect scalar structure, we need to define $\apar$
and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$
(merging of coeffect annotations). The $\azip$ operation (corresponding to $\czip$
in flat coeffect calculus) is provided explicitly.

@@

\begin{lemma}
\label{thm:further-weak}
A unified coeffect calculus capturing flat properties admits weakening that does not change the
coeffect annotation.
\end{lemma}
\begin{proof}
The rule is admissible as shown using the following derivation:
\begin{equation*}
\inference
  { \inference
      {\coctx{\Gamma}{ \cclrd{r} } \vdash e : \tau}
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \;\apar\; \alift{ \czero }} \vdash e : \tau} }
  { \inference
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \,\cpar\, \czero} \vdash e : \tau}
      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r}} \vdash e : \tau} }
\end{equation*}

@@

\begin{lemma}
A unified coeffect calculus capturing flat properties admits exchange rule that does not change the
coeffect annotation.
\end{lemma}
\begin{proof}
We use the idempotence of $\czip$ and $\cpar$ together with the (\emph{exch}) rule:
\begin{equation*}
\inference
  { \inference
      { \inference
          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r}} \vdash e : \tau}
          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r}} \vdash e : \tau} }
      { \coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}} \vdash e : \tau } }
  { \inference
      {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \apar \alift{\cclrd{r}} \apar \alift{\cclrd{r}} \apar \cclrd{r}} \vdash e : \tau}
      {\inference
         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}} \vdash e : \tau}
         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}} \vdash e : \tau} } }
\end{equation*}

@@

\noindent
Although the different readings do not affect formal properties of the systems, it is important
to understand the difference when discussing the two systems as they provide a different
intuition.
